Mapping short reads, faithfully

### 2020-02-10 manuscript.tex
Make minor modifications in the text

% Enter dates of publication
\copyrightyear{2020}
\pubdate{1 January 2020}
\pubyear{2020}
\jvolume{37}
\jissue{12}

@@

\author{%
Eduard Valera Zorita\,$^{1,\text{\#}}$,
Ruggero Cortini\,$^{1,\text{\#}}$,
Guillaume J. Filion\,$^{1,2,3}$
\footnote{To whom correspondence should be addressed.
Email: guillaume.filion@gmail.com}}

\address{%
$^{1}$Center for Genomic Regulation (CRG), The Barcelona Institute
of Science and Technology, Dr. Aiguader 88, Barcelona 08003, Spain;
$^{2}$University Pompeu Fabra (UPF), Barcelona, Spain;
$^{3}$present address: Department of Biological Sciences, University
of Toronto Scarborough, Toronto, ON, Canada; $^{\text{\#}}$ equal
contributions.}
% Affiliation must include:
% Department name, institution name, full road and district address,
% state, Zip or postal code, country

@@

\begin{abstract}
Mapping is the process of finding the original location of a DNA read in a
reference sequence, typically a genome. Short read mappers are used in
most applications of high-throughput sequencing. As such, they must be
continuously improved to keep up with increasing needs. To gain speed,
mappers rely on a heuristic known as seeding, which has a small
probability of causing mapping errors. For lack of a method to compute
such probabilities, mappers have so far used \textit{ad hoc}
approximations of variable quality. Here we devise the elements of a
strategy to map short reads faithfully, \textit{i.e.}, to provide accurate
probabilities of mapping errors. The key is to estimate the number of
paralogus of the target so as to spend more time on the reads that can be
mapped with high confidence. This strategy highlights the existence of a
class of reads that can be mapped with unprecedented confidence. We
implement this strategy in a prototype mapper that is competitive with
state-of-the-art mappers BWA-MEM and Bowtie2, with the benefit of
faithfulness. The software is open-source and availble from
\url{https://github.com/gui11aume/smmfdp}.
\end{abstract}

@@

The importance of faithfulness is recognized in the specifications of the
SAM format \cite{pmid19505943}, featuring a mandatory \emph{mapping
quality} field MAPQ, defined as ``$-10 \cdot \log_{10}Pr$\{mapping
position is wrong\}''. However, there is presently no way to estimate this
probability, and therefore no way to respect the standard. Some mappers
like STAR~\cite{pmid23104886} use the MAPQ field as a qualitative scale.
Some others like GEM~\cite{pmid23103880} do not provide a mapping quality.
As for the mappers aiming to comply with the SAM standard, the scores are
undocumented, variable between mappers and versions, often spreading
puzzlement and frustration in the community~\cite{ACGT, QCFail}.

@@

Assuming that the read is mapped with $x$ mismatches, the probability that
the location is incorrect is computed as
\begin{align}
\label{eq_bayes}
P(N \geq 1) \cdot \frac{P_s}{P_s + (1-P_s)e^{\min(a-b,1)}},
\end{align}
where $P_s$ is the prior probability provided by \texttt{sesame}, where
$a-b=x \cdot \log(p/\lambda)+(L-x) \cdot \log((1-p)/(1-\lambda))$, and
where $P(N \geq 1)$ is a probability computed during the estimation of $N$ (see
the Supplementary Material for more detail).

@@

@@

The key in the flow chart of Figure~\ref{fig_tree} is to dispatch the
reads based on $N$, the number of paralogs of the target. Estimating $N$
must be fast so that this step does not become a bottleneck. To achieve
this, we propose to use the seeding process itself. MEM seeds are best
computed using the backward search~\cite{ferragina2000opportunistic}, an
algorithm that returns the number of hits in the genome as the query is
extended backward. If the target has no or few paralogs
(Figure~\ref{fig_back}a), the number of hits will quickly go below 2 (when
there is only one hit, it is most likely the target itself). In contrast,
if the target has many paralogs in the genome (Figure~\ref{fig_back}b),
the number of hits will remain at 2 or above for may iterations.

@@

In terms of speed, MMP is 2--4 faster than BWA-MEM and Bowtie2 on the
present data set (Figure~\ref{fig_timem}a) but the memory footprint is
approximately 2--3 times higher (Figure~\ref{fig_timem}b), mostly because
the reverse complement of the genome is added to the index.

@@

Here we designed and implemented a strategy to map short reads faithfully.
The principles are based on key insight gained from models to compute
seeding probabilities~\cite{Filion619155}, showing that MEM seeds are
efficient only when the target has few paralogues (Figure~\ref{fig_MEM}).
This motivates the design of a strategy where we first estimate the copy
number of the target (Figure~\ref{fig_back}) to dispatch reads to
different mapping subroutines (Figure~\ref{fig_tree}). This process
revealed the existence of super reads, \textit{i.e.}, reads that can be
mapped with extremely high confidence (Figures~\ref{fig_supertest} and
\ref{fig_supread}), together setting the basis of an efficient mapping
strategy. We wrote a mapper based on this strategy and we showed that it
is competitive with BWA-MEM and Bowtie (Figures~\ref{fig_faith},
\ref{fig_acc} and \ref{fig_timem}). Importantly, the mapping process
itself relies only on standard algorithm, so all the benefits of the
mapper come from faithfulness. Overall, this demonstrates that improving
faithfulness is a fruitful strategy to improve short-read mappers.

@@

We would like to thank Santiago Marco Sola, Catalina Romero and Nicholas
Stroustrup for their critical comments on the manuscript.
This work was funded by the Spanish Ministry of Economy, Industry and
Competitiveness (MEIC, Plan Estatal PGC2018-099807-B-I00) and by the
European Research Council (Synergy Grant 609989). R.~C. was supported by
the People Programme (Marie Curie Actions) of the European Union's Seventh
Framework Programme (FP7/2007-2013) under REA grant agreement 608959. We
acknowledge the financial support of the MEIC to the EMBL partnership, the
Centro de Excelencia Severo Ochoa and the CERCA Programme / Generalitat de
Catalunya.

@@

\clearpage

\subsection{Supplementary Material}
\subsubsection{Computing $P(N \geq 1)$.}
If the target is unique, the read is aligned against a random sequence
whenever the mapping location is incorrect. In this case, it is easy to
realize that the genomic sequence has no homology for the read. In
practice we can consider that reads cannot be mapped incorrectly when the
target is unique (they can be unmapped though). In other words, mapping
errors are possible only if the target has at least one paralog,
\textit{i.e.} if $N \geq 1$. As a consequence, the probability of a
mapping error is proportial to $P(N \geq 1)$. This term is implicit in the
mapping quality of super reads (\ref{eq_super}), and set to 1 for
low-quality mapping in expression (\ref{eq_low}). In expressions
(\ref{eq_bayes}) and (\ref{eq_x0}), it must be computed explicitly.

The estimate is based on the method shown in Figure~\ref{fig_back}, where
we use the end of the read with the shortest query. We consider three
cases: $i.$ the target has a paralog, $ii.$ only one end of the target has
a paralog, and $iii.$ the target is unique. Case $ii.$ can occur when the
read straddles the end of a transposon or other repeated sequence. We
consider that $N \geq 1$ only in case $i.$ where the sequence has a
full-length paralog.

\clearpage

\begin{figure}
\begin{center}
\includegraphics[scale=.53]{timem_supp.pdf}
\end{center}
\caption{Resource usage at 2, 5 and 10\% error rate. The representations
are otherwise as in Figure~\ref{fig_timem} (the memory footprint does not
depend on the error rate of the sequencing process).}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[scale=.54]{accQ2_fig.pdf}
\end{center}
\caption{Mapping accuracy with 2\% error rate. The representations are
otherwise as in Figure~\ref{fig_acc}.}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[scale=.54]{accQ5_fig.pdf}
\end{center}
\caption{Mapping accuracy with 5\% error rate. The representations are
otherwise as in Figure~\ref{fig_acc}.}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[scale=.54]{accQ10_fig.pdf}
\end{center}
\caption{Mapping accuracy with 10\% error rate. The representations are
otherwise as in Figure~\ref{fig_acc}.}
\end{figure}

\clearpage

\begin{figure*}[t]
\begin{center}
\includegraphics[scale=.24]{faith_Q2.pdf}
\end{center}
\caption{Faithfulness with 2\% error rate. The representations are
otherwise the same as in Figure~\ref{fig_faith}.}
\end{figure*}

\begin{figure*}[t]
\begin{center}
\includegraphics[scale=.24]{faith_Q5.pdf}
\end{center}
\caption{Faithfulness with 5\% error rate. The representations are
otherwise the same as in Figure~\ref{fig_faith}.}
\end{figure*}

\begin{figure*}[t]
\begin{center}
\includegraphics[scale=.24]{faith_Q10.pdf}
\end{center}
\caption{Faithfulness with 10\% error rate. The representations are
otherwise the same as in Figure~\ref{fig_faith}.}
\end{figure*}

>>>>

% Enter dates of publication
\copyrightyear{2020}
\pubdate{31 December 2020}
\pubyear{2020}
\jvolume{XX}
\jissue{YY}

@@

\author{%
Eduard Valera Zorita\,$^1$,
Ruggero Cortini\,$^1$,
Guillaume J. Filion\,$^{1,2,3}$
\footnote{To whom correspondence should be addressed.
Email: guillaume.filion@gmail.com}}

\address{%
$^{1}$Center for Genomic Regulation (CRG), The Barcelona Institute
of Science and Technology, Dr. Aiguader 88, Barcelona 08003, Spain;
$^{2}$University Pompeu Fabra (UPF), Barcelona, Spain;
$^{3}$present address: Department of Biological Sciences, University
of Toronto Scarborough, Toronto, ON, Canada.}
% Affiliation must include:
% Department name, institution name, full road and district address,
% state, Zip or postal code, country

@@

\begin{abstract}
Mapping is the process of finding the original location of a DNA read in a
reference sequence, typically a genome. Short read mappers are used in
most applications of high-throughput sequencing. As such, they must be
continuously improved to keep up with increasing needs. To gain speed,
many mappers rely on seeding heuristics that can negatively affect the
mapping accuracy. For lack of a method to compute such probabilities,
mappers have so far used \textit{ad hoc} approximations of variable
quality. Here we devise the elements of a strategy to map short reads
faithfully, \textit{i.e.}, to provide accurate probabilities of mapping
errors. The key is to estimate the number of paralogus of the target so as
to spend more time on the reads that can be mapped with high confidence.
This strategy highlights the existence of a class of reads that can be
mapped with unprecedented confidence. We implement this strategy in a
prototype mapper that is competitive with state-of-the-art mappers BWA-MEM
and Bowtie2, with the benefit of faithfulness. The software is open-source
and availble from \url{https://github.com/gui11aume/smmfdp}.
\end{abstract}

@@

The importance of faithfulness is recognized in the specifications of the
SAM format \cite{pmid19505943}, featuring a mandatory \emph{mapping
quality} field MAPQ, defined as ``$-10 \cdot \log_{10}Pr$\{mapping
position is wrong\}''. However, the proposed methods to estimate this
probability are accurate only for some genomes and some sequencing
technologies.
THIS NEEDS TO BE WRITTEN BETTER.
%However, there is presently no way to estimate this
%probability, and therefore no way to respect the standard.
Some mappers
like STAR~\cite{pmid23104886} use the MAPQ field as a qualitative scale.
%Some others like GEM~\cite{pmid23103880} do not provide a mapping quality.
As for the mappers aiming to comply with the SAM standard, the scores are
undocumented, variable between mappers and versions, often spreading
puzzlement and frustration in the community~\cite{ACGT, QCFail}.

@@

Assuming that the read is mapped with $x$ mismatches, the probability that
the location is incorrect is computed as
\begin{align}
\label{eq_bayes}
P(N \geq 1) \cdot \frac{P_s}{P_s + (1-P_s)e^{\min(a-b,1)}},
\end{align}
where $P_s$ is the prior probability provided by \texttt{sesame}, where
$a-b=x \cdot \log(p/\lambda)+(L-x) \cdot \log((1-p)/(1-\lambda))$, and
where $P(N \geq 1)$ is a probability computed during the estimation of $N$
(see below).

@@

\subsubsection{Estimating $P(N \geq 1)$.} This quantity appears in
expressions (\ref{eq_bayes}) and (\ref{eq_x0}). The estimate is based on
the method shown in Figure~\ref{fig_back}, where we use the end of the
read with the shortest query. We consider three cases: $i.$ the target has
a paralog, $ii.$ only one end of the target has a paralog, and $iii.$ the
target is unique. Case $ii.$ can occur when the read straddles the end of
a transposon or other repeated sequence. We consider that $N \geq 1$ only
in case $i.$ where the sequence has a full-length paralog.

@@

The key in the flow chart of Figure~\ref{fig_tree} is to dispatch the
reads based the number $N$ of paralogs of the target, which is an
indirect measure of the mappability of the read~\cite{pmid22276185}.
Estimating $N$ must be fast so that this step does not become a
bottleneck. To achieve this, we propose to use the seeding process itself.
MEM seeds are best computed using the backward
search~\cite{ferragina2000opportunistic}, an algorithm that returns the
number of hits in the genome as the query is extended backward. If the
target has no or few paralogs (Figure~\ref{fig_back}a), the number of hits
will quickly go below 2 (when there is only one hit, it is most likely the
target itself). In contrast, if the target has many paralogs in the genome
(Figure~\ref{fig_back}b), the number of hits will remain at 2 or above for
may iterations.

@@

In terms of speed, MMP is 2--4 faster than BWA-MEM and Bowtie2 on the
present data set (Figure~\ref{fig_timem}a), with a memory footprint that
is approximately 2--3 times higher (Figure~\ref{fig_timem}b). Part of the
speed up is due to the implementation and the larger stress on memory, but
the strategy highlighted in Figure~\ref{fig_tree} aims to reduce the
number of alignments when reads cannot be mapped accurately. For instance,
MMP spends 8.3\% of the time on incorrectly mapped reads, versus 29.2\%
and 12.5\% for BWA-MEM and Bowtie, respectively (100 nucleotide reads in
the human genome, data not shown).

@@

Here we designed and implemented a strategy to map short reads faithfully.
The principles are based on key insight gained from models to compute
seeding probabilities~\cite{Filion619155}, showing that MEM seeds are
efficient only when the target has few paralogues (Figure~\ref{fig_MEM}).
This motivates the design of a strategy where we first estimate the copy
number of the target (Figure~\ref{fig_back}) to dispatch reads to
different mapping subroutines (Figure~\ref{fig_tree}). This amounts to
allocating more resources to the most mappable reads~\cite{pmid22276185}.
This process revealed the existence of super reads, \textit{i.e.}, reads
that can be mapped with extremely high confidence
(Figures~\ref{fig_supertest} and \ref{fig_supread}), together setting the
basis of an efficient mapping strategy. We wrote a mapper based on this
strategy and we showed that it is competitive with BWA-MEM and Bowtie
(Figures~\ref{fig_faith}, \ref{fig_acc} and \ref{fig_timem}). Importantly,
the mapping process itself relies only on standard algorithm, so all the
benefits of the mapper come from faithfulness. Overall, this demonstrates
that improving faithfulness is a fruitful strategy to improve short-read
mappers.

@@

We would like to thank Santiago Marco-Sola, Catalina Romero and Nicholas
Stroustrup for their critical comments on the manuscript.
This work was funded by the Spanish Ministry of Economy, Industry and
Competitiveness (MEIC, Plan Estatal PGC2018-099807-B-I00) and by the
European Research Council (Synergy Grant 609989). R.~C. was supported by
the People Programme (Marie Curie Actions) of the European Union's Seventh
Framework Programme (FP7/2007-2013) under REA grant agreement 608959. We
acknowledge the financial support of the MEIC to the EMBL partnership, the
Centro de Excelencia Severo Ochoa and the CERCA Programme / Generalitat de
Catalunya.

@@
